Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-26 ind_files/04_field_eqns.tex
Add some references for field eqns/SPDEs

So far, we have dealt with populations in which individuals come in countably many different kinds. While developing these models, we have been on mathematically solid ground that is well understood by statistical physicists and mathematicians. However, things become more complicated when we deal with `quantitative' traits. Traits like body size, body weight, or beak length, often take on uncountably many values (say, all values in the interval $[0,1]$, for example). In this case, we cannot describe the population using a vector as we did before, but instead require a function. More precisely, if the set of all possible trait values is $\mathcal{T}$, we will characterize the population using a special kind of function $\phi^{(t)}$ such that the quantity $\int_{A}\phi^{(t)}(x)dx$ gives us the number of individuals that are in any `nice' region $A \subset \mathcal{T}$ of the possible trait space\footnote{The mathematically informed reader may notice that this sounds like I am trying to dance around the word `measure'. Indeed, we are really looking to construct branching processes that take values in some nice space of measures that can be endowed with sufficient mathematical structure for notions like convergence and integration to make sense. All the Dirac deltas that will turn up shortly are `properly' viewed as measures, integrals with Dirac deltas in the integrand are `properly' interpreted as integration with respect to the Dirac measure, and the functional derivative is to be `properly' interpreted as a Gateaux or  Fr\'echet derivative assuming that the relevant function space has enough structure for these notions to make sense. If one tries to be careful about these things, they will quickly find themselves drowning in a quagmire of mathematical formalism.
% For example, one can't really justify using informal tools like the functional derivative, which is ill-defined and `properly' interpreted as a Fr\'echet derivative in a suitable normed function space (making the use in conjunction with Dirac measures/distributions somewhat dubious without further elaboration on the space and norm in question). The existence of a probability density function for this process is also \emph{a priori} somewhat questionable and requires detailed arguments using tools like the Radon-Nikodym theorem. Here, we adopt the physicist's solution of ignoring all such potential difficulties and assuming a clever mathematician can make sure everything works out using arcane tools such as ``functional analysis" and  ``distribution theory".
If you know and care about enough mathematics for this to really bother you, see~\citet{champagnat_unifying_2006} or~\citet{champagnat_individual_2008} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.} at time $t$. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$.

@@

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
We will assume that our process is nice enough that the hypotheses of the Radon-Nikodym theorem are satisfied and we can find a density function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by $\int_{\mathcal{T}}P(\nu,t)dx$. In one dimension, this is not necessarily a very big ask for biologically reasonable choices of the birth and death functionals~(\cite{dawson_stochastic_1975,walsh_introduction_1986,dawson_stochastic_2000}; Also see proposition 3.1 in~\cite{champagnat_individual_2008}).

@@

Equation \eqref{cts_replicator_mutator} also recovers~\posscite{kimura_stochastic_1965} continuum-of-alleles model when each $x$ is viewed as an allele, $Q(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $Q(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m:\mathbb{R} \to [0,\infty)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that I have implicitly been assuming that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb
R}\psi(x,t)dx < \infty \ \forall \ t$. Thus, for any fixed $t > 0$, we have $m(y-z)\psi(z,t) \in  \mathcal{L}^{1}(\mathbb{R}\times\mathbb{R})$ and we can use the Fubini-Tonnelli theorem to interchange the order of integration of iterated integrals of $m(y-z)\psi(y)$. We are now ready to evaluate the rightmost integral of \eqref{cts_replicator_mutator}.

@@

can be thought of now as the functional analogue of a directional derivative of $\mathcal{A}^-(x|\psi)$ in the direction of the function $\zeta$.

>>>>

So far, we have dealt with populations in which individuals come in countably many different kinds. While developing these models, we have been on mathematically solid ground that is well understood by statistical physicists and mathematicians. However, things become more complicated when we deal with `quantitative' traits. Traits like body size, body weight, or beak length, often take on uncountably many values (say, all values in the interval $[0,1]$, for example). In this case, we cannot describe the population using a vector as we did before, but instead require a function. More precisely, if the set of all possible trait values is $\mathcal{T}$, we will characterize the population using a special kind of function $\phi^{(t)}$ such that the quantity $\int_{A}\phi^{(t)}(x)dx$ gives us the number of individuals that are in any `nice' region $A \subset \mathcal{T}$ of the possible trait space\footnote{The mathematically informed reader may notice that this sounds like I am trying to dance around the word `measure'. Indeed, we are really looking to construct branching processes that take values in some nice space of measures that can be endowed with sufficient mathematical structure for notions like convergence and integration to make sense and are absolutely continuous with respect to the Lebesgue measure. All the Dirac deltas that will turn up shortly are `properly' viewed as measures, integrals with Dirac deltas in the integrand are `properly' interpreted as integration with respect to the Dirac measure, and the functional derivative is to be `properly' interpreted as a Gateaux or  Fr\'echet derivative assuming that the relevant function space has enough structure for these notions to make sense. If one tries to be careful about these things, they will quickly find themselves drowning in a quagmire of mathematical formalism.
% For example, one can't really justify using informal tools like the functional derivative, which is ill-defined and `properly' interpreted as a Fr\'echet derivative in a suitable normed function space (making the use in conjunction with Dirac measures/distributions somewhat dubious without further elaboration on the space and norm in question). The existence of a probability density function for this process is also \emph{a priori} somewhat questionable and requires detailed arguments using tools like the Radon-Nikodym theorem. Here, we adopt the physicist's solution of ignoring all such potential difficulties and assuming a clever mathematician can make sure everything works out using arcane tools such as ``functional analysis" and  ``distribution theory".
If you know and care about enough mathematics for this to really bother you, see~\citet{champagnat_unifying_2006} or~\citet{champagnat_individual_2008} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.} at time $t$. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$.

@@

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
We will assume that our process is nice enough that the hypotheses of the Radon-Nikodym theorem are satisfied and we can find a density function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by $\int_{\mathcal{T}}P(\nu,t)dx$. In one dimension, this is not necessarily a difficult requirement for biologically reasonable choices of the birth and death functionals~(\cite{dawson_stochastic_1975,walsh_introduction_1986,konno_stochastic_1988,reimers_one_1989,dawson_stochastic_2000}; Also see proposition 3.1 in~\cite{champagnat_individual_2008}).

@@

Equation \eqref{cts_replicator_mutator} also recovers~\posscite{kimura_stochastic_1965} continuum-of-alleles model when each $x$ is viewed as an allele, $Q(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $Q(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m:\mathbb{R} \to [0,\infty)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that I have implicitly been assuming\footnote{This actually need not be an assumption, and can be derived as a theorem under biologically reasonable conditions on the birth/death functionals upon taking the rigorous approach to both the system size expansion (rescaling) and the infinite population limit. See, for example,~\citet{etheridge_introduction_2000} or proposition 4.1 in~\citet{champagnat_individual_2008}} that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb
R}\psi(x,t)dx < \infty \ \forall \ t$. Thus, for any fixed $t > 0$, we have $m(y-z)\psi(z,t) \in  \mathcal{L}^{1}(\mathbb{R}\times\mathbb{R})$ and we can use the Fubini-Tonnelli theorem to interchange the order of integration of iterated integrals of $m(y-z)\psi(y)$. We are now ready to evaluate the rightmost integral of \eqref{cts_replicator_mutator}.

@@

can be thought of now as the functional analogue of a directional derivative of $\mathcal{A}^-(x|\psi)$ in the direction of the `fluctuation' $\zeta$, now a function. In Appendix \ref{App_examples}, we will see how this linear approximation can be used to study phenotypic clustering/evolutionary branching using spectral methods first introduced for various specific models by~\citet{rogers_demographic_2012} and further expanded upon in~\citet{rogers_modes_2015}.

### 2023-03-26 ind_files/05_unification_discussion.tex
Add some references for field eqns/SPDEs

Currently, (stochastic) field equations are primarily used by physicists working in areas such as statistical field theory and quantum field theory, and are largely attacked using ingenious and often somewhat system-specific tools that may not necessarily be mathematically well understood~(\cite{carmona_stochastic_1999}; but see \cite{bogachev_fokker-planck-kolmogorov_2015} for a rigorous treatment of some infinite-dimensional Fokker-Planck equations). Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using some of these tools - For example, a careful reading of the literature provides many hints (albeit couched in physics language) for how (quantum) field theoretic ideas such as the Fock space representation~\citep{del_razo_probabilistic_2022}, `loop diagrams/expansions'~\citep{hochberg_effective_1999,dodd_many-body_2009}, and the often closely related path integral formalism~\citep{doi_second_1976, peliti_path_1985,hochberg_effective_1999, chow_path_2015, weber_master_2017}, can all be co-opted to coax biological insights from the processes modeled by equation \eqref{disc_functional_field_eqns}.

Equation \eqref{disc_functional_field_eqns} thus provides a (relatively) more accessible alternative formulation that does not (explicitly) rely on tools from relatively `advanced' mathematical fields such as martingale theory and stochastic analysis that are typically used to study the evolution of quantitative traits in finite, fluctuating populations~\citep{dawson_stochastic_1975,fleming_measure-valued_1979,ethier_markov_1986,champagnat_unifying_2006,etheridge_mathematical_2011,week_white_2021}. This is intended to make the models and ideas more accessible to theorists without formal training in mathematical areas such as measure theory and functional analysis and encourage the use of techniques from physics and related domains.

Importantly, the formalism I develop here in terms of functional Fokker-Planck equations likely does \emph{not} carry over to the study of higher dimensional quantitative traits (or populations which vary in two or more one-dimensional quantitative traits) because these processes are routinely badly behaved in higher dimensions: In particular, a probability density $P(\phi,t)$ with respect to the reals frequently does not even exist in higher dimensions if one has any biologically non-trivial features such as interactions between types (\cite{fleming_measure-valued_1979, walsh_introduction_1986}; Also see for example~\cite{evans_measure-valued_1994}), rendering equation \eqref{disc_functional_field_eqns} entirely meaningless. My (admittedly limited) understanding is that such processes are also rather difficult to study in two or more dimensions from the SPDE side for similar reasons since one is forced to work with distribution valued (rather than function valued) solutions and thus needs to use a considerable amount of advanced functional analysis for even existence/uniqueness results~\citep{walsh_introduction_1986,carmona_stochastic_1999,prevot_concise_2007,liu_stochastic_2015,balan_gentle_2018}. It may well be the case that concrete biologically useful progress in this direction requires new mathematics altogether, a situation increasingly also encountered in other areas of mathematical biology~\citep{vittadello_open_2022}. Just as theoretical physics has done in the past, the pursuit of general mathematical descriptions of biological theory may therefore also inspire new ideas that are of independent interest to `pure' mathematicians~\citep{cohen_mathematics_2004}.

>>>>

Currently, (stochastic) field equations are primarily used by physicists working in areas such as statistical field theory and quantum field theory, and are largely attacked using ingenious and often somewhat system-specific tools that may not necessarily be mathematically well understood~(\cite{carmona_stochastic_1999}; but see \cite{bogachev_fokker-planck-kolmogorov_2015} for a rigorous treatment of some infinite-dimensional Fokker-Planck equations). Equation \eqref{disc_functional_field_eqns} opens up the study of quantitative trait dynamics in finite fluctuating populations to analysis using some of these tools - For example, a careful reading of the literature provides many hints (albeit couched in physics language) for how (quantum) field theoretic ideas such as the Fock space representation~\citep{del_razo_probabilistic_2022}, `loop diagrams/expansions'~\citep{hochberg_effective_1999,dodd_many-body_2009}, and the often closely related path integral formalism~\citep{doi_second_1976, peliti_path_1985,hochberg_effective_1999, chow_path_2015, weber_master_2017}, can all be co-opted to coax biological insights from the processes modeled by equation \eqref{disc_functional_field_eqns}. The formalism developed in Chapter \ref{chap_infD_processes} is intended to encourage the use of such techniques and make some of the models and ideas more accessible to theorists without formal training in mathematical areas such as measure theory and functional analysis.

Importantly, the formalism I develop here in terms of functional Fokker-Planck equations likely does \emph{not} carry over to the study of higher dimensional quantitative traits (or populations which vary in two or more one-dimensional quantitative traits) because these processes are routinely badly behaved in higher dimensions: In particular, a probability density $P(\phi,t)$ with respect to the reals frequently does not even exist in higher dimensions if one has any biologically non-trivial features such as interactions between types (\cite{fleming_measure-valued_1979, walsh_introduction_1986,etheridge_introduction_2000}; Also see for example~\cite{evans_measure-valued_1994}), rendering equation \eqref{disc_functional_field_eqns} entirely meaningless. My (admittedly limited) understanding is that such processes are also rather difficult to study in two or more dimensions from the SPDE side for similar reasons~\citep{etheridge_introduction_2000}, where the lack of a density leads to distribution-valued (rather than function-valued) solutions at best and thus demands a considerable amount of advanced functional analysis to establish even `basic' existence/uniqueness results~\citep{walsh_introduction_1986,carmona_stochastic_1999,prevot_concise_2007,liu_stochastic_2015,balan_gentle_2018}. It may well be the case that concrete biologically useful progress in this direction requires new mathematics altogether, a situation increasingly also encountered in other areas of mathematical biology~\citep{vittadello_open_2022}. Just as theoretical physics has done in the past, the pursuit of general mathematical descriptions of biological theory may therefore also inspire new ideas that are of independent interest to `pure' mathematicians~\citep{cohen_mathematics_2004}.
